A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-07-14 paper/fascd.tex
address all PF comments re Results section

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}, allowing $P_1$ or $Q_1$ elements and based on standard nested mesh refinement.  Here we show numerical results for the example problems presented in Section \ref{sec:vi}, with a focus on the scaling of FAScd iterations as the mesh resolution increases.  Our main observations are that each V-cycle dramatically reduces the residual norm, and that very few F-cycles solve most examples to within discretization error.

\begin{example} \label{ex:results:classical}
We consider the unilateral 2D Laplacian obstacle problem presented in Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\} \subset W^{1,2}(\Omega)$.  We consider two problems with very different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$, both over square $\Omega$.   The standard ``ball'' problem has an upper hemisphere obstacle $\psi$ \cite[Chapter 12]{Bueler2021}.  In the ``spiral'' problem \cite[Example 7.1.1]{GraeserKornhuber2009}, $\psi$ is constructed to produce a more interesting coincidence set; see Figure \ref{fig:results:classical}.

For this experiment we used a coarsest mesh of $m_0=5^2$ vertices (32 triangles).  The finest mesh had $m_9=2049^2 \approx 4.2 \times 10^{6}$ vertices.  As our smoother we applied a single RS Newton iteration (Section \ref{sec:implementation}), with three (fixed) IC-preconditioned CG iterations to approximately solve the arising linear systems.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  The initial iterate was set to $w=\max\{0,\psi\}$.

Table \ref{tab:results:classical} shows the resulting iteration counts for finest meshes of $m_J$ vertices.  When the RS Newton smoother is applied by itself as a solver (``RS only'') it shows rapidly-increasing iterations as expected, roughly proportional to resolution.  By contrast, V-cycle iterations increase slowly, perhaps logarithmically with the resolution.  Noting that for F-cycles we are counting the number of V-cycles after the ramp (Section \ref{sec:implementation}), we see that a constant number of cycles solves the problem.  Comparing the ball and spiral results, the complexity of the coincidence set or free boundary does not appear to strongly influence FAScd performance \pef{unlike \cite{GraeserKornhuber2009}?}. Each V-cycle does the work of about eight finest-level preconditioned Krylov (CG+IC) iterations; for the F-cycle results, where three iterations suffice at all resolutions, we see close to textbook multigrid efficiency\footnote{Textbook multigrid efficiency is imperfectly defined, but consider \cite{BrownSmithAhmadia2013}: ``an order of magnitude reduction in residual per iteration and solution of the fine-level system at a small multiple of the cost of a residual evaluation.''  Work of less than ten residual evaluations is often mentioned as the standard.} despite the low solution regularity.
\end{example}

@@

\begin{example}  \label{ex:results:plap}
We again consider Example \ref{ex:plaplacian} with a unilateral constraint, but now for the nonlinear $p$-Laplacian. We investigate both fast ($p=1.5$) and degenerate ($p=4.0$) diffusivities.\footnote{``Fast'' and ``degenerate'' refer to $|\grad u|^{p-2}\to \infty$ and $\to 0$, respectively; ellipticity is lost in the latter case.  Well-posedness and ($p$-dependent) solution regularity for $p$-Laplacian obstacle problems is considered in \cite{ChoeLewis1991}.} We consider the problem in one dimension for ease of visualization. This example reveals the sometimes strong contrast between $V(1,0)$ and $V(0,1)$ results (\pr{up}, \pr{down} $=0$ in Algorithm \ref{alg:fascd}, respectively), and it emphasizes the need for robust smoothers.

Let $\Omega=(-3,3)$, $\psi(x) = -0.2|x|$, $\mathcal{K} = \{v \ge \psi\} \subset W^{1,p}(\Omega)$, and $\ell(v) = \int_\Omega g v$ where $g(x)=1$ for $|x|<1$ and $g(x)=-1$ for $|x|>1$.  The exact solution of VI \eqref{eq:vi} is easily computed for any $p$; see Figure \ref{fig:results:plap}. \pef{Write down the exact solution?}  For this experiment we used a coarsest mesh of $m_0=7$ vertices and a finest mesh with $m_9=3073$.  The default smoother uses three (fixed) iterations of RS Newton, with LU solution of the arising linear systems.\footnote{This is an optimal $O(m_j)$ smoother in 1D.}  The coarse solver is the same but iterated to convergence.  The initial iterate was $w=\psi$, and \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$.  Figure \ref{fig:imagesvcycle} illustrates the major steps of an example V-cycle, showing each of the coarse-correction VI subproblems and solutions.

@@

Solvers implemented in Firedrake can exploit parallel solvers from the PETSc library \cite{Balayetal2023}.  The above Examples, run in serial so as to fix the smoother,\footnote{The action of parallel preconditioners can depend on the number of processes \cite[for example]{Bueler2021}.} also run efficiently in parallel.  The next Example applies a parallel solver, and Example \ref{ex:results:sia} demonstrates the parallel weak scaling \cite{Bueler2021} of FAScd.

\begin{example}  \label{ex:results:advdiff}
We consider the advection-diffusion VI problem in Example \ref{ex:advectiondiffusion}.  Let $\Omega=(-1,1)^d$ for $d=2,3$, the diffusivity $\eps=0.1$ (equation \eqref{eq:advectiondiffusionstrong}), and we impose zero Dirichlet boundary conditions.  The 2D problem satisfied the coercivity hypothesis with a divergence-free velocity field $\bX = (7+5y,-5x)$, but in 3D we set $\bX = (7+5y,-5x,2z)$ so $\grad\cdot\bX \ne 0$. \pef{Why?} The discontinuous source term $\phi \in L^\infty(\Omega)$ satisfied $\phi\ge 0$ in half of the domain, with positive circular patches, while values were negative in the other half.  \pef{State where to find the full specification of the source term.} While the other Examples in this Section are unilateral, here the solution was box-constrained: $0 \le u \le 1$.  Figure \ref{fig:results:advdiff} shows coincidence sets for the 2D problem; the 3D sets are comparable in regularity and general pattern.

@@

F-cycle iteration results are shown in Table \ref{tab:results:advdiff}.  Despite the hint of growing iterations under mesh refinement, a few iterations sufficed up to memory-limited resolutions.  For any of the finer mesh runs here, the FAScd F-cycle solver took less time than generating the $P_1$ mesh hierarchy.  The local Péclet number $\text{Pe}=\|\bX(x,y,z)\|_2 L \eps^{-1}$, with $L=2$ the width of $\Omega$, varied from 40 to 260.  Thus the $m_J\le 121^d$ problems were significantly advection-dominated while those with $m_J\ge 481^d$ were diffusion-dominated.  \pef{I would have said you need $Pe < 1$ to be diffusion-dominated, not have $Pe = 40$?} Nonetheless the smoother was the same and F-cycle iterations were nearly constant.
\end{example}

% see git@bitbucket.org:pefarrell/fascd.git: examples/pollutant.py
% and https://github.com/bueler/mcd-extended: paper/results/pollutant.sh|txt
\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}c@{\hskip 4mm}c}
\emph{levels} & $m_J$ & F ($d=2$) & F ($d=3$) \\ \cmidrule{1-4}
 1 &    $16^d$ & 1 & 1 \\
 2 &    $31^d$ & 2 & 2 \\
 3 &    $61^d$ & 2 & 2 \\
 4 &   $121^d$ & 2 & 2 \\
 5 &   $241^d$ & 2 & 3 \\
 6 &   $481^d$ & 2 \\
 7 &   $961^d$ & 3 \\
 8 &  $1921^d$ & 3 \\
 9 &  $3841^d$ & 3
\end{tabular}
\bigskip
\caption{F-cycle iterations for 2D ($P_1$, serial) and 3D ($Q_1$, 16 processes) solutions of the advection-diffusion bi-obstacle VI problem in Example \ref{ex:results:advdiff}.}
\label{tab:results:advdiff}
\end{table}

\begin{example}   \label{ex:results:sia}
We consider the steady shallow ice approximation (SIA) obstacle problem presented in Example \ref{ex:sia}, in which the surface elevation $s$ is constrained to exceed the bedrock elevation $b$.  We set $\Omega=(0,L)^2$ with $L=1800$ km.  The data consist of a radial surface mass balance function $a$ \cite[equation (5.122)]{GreveBlatter2009} and a bumpy, synthetic map $b\in C^1(\Omega)$, with an elevation range of $1800$ m.  A numerical solution is shown in Figure \ref{fig:results:siascene}, colored by the solution surface flow speed, i.e.~the magnitude of $\mathbf{u} = (n+1)(n+2)^{-1} \Gamma (s-b)^{n+1} |\grad s|^{n-1} \grad s$.  As noted in Example \ref{ex:sia}, and seen in this Figure, $|\grad s|$ is unbounded at the free boundary; the solution $s$, while continuous, has very low regularity.

@@

We applied FAScd F-cycles using $Q_1$ elements and a smoother of four (fixed) iterations of RS Newton (Section \ref{sec:implementation}), with quadratic backtracking line search.  The step equations were approximately solved, in parallel, by three (fixed) iterations of preconditioned GMRES, again with a ASM+ILU preconditioner.  The coarse solver was RS Newton with LU applied redundantly over all processors.  The coarse mesh was relatively fine, with $m_0=21^2$ vertices and 400 cells.  Because of the low regularity of the solution we used \id{rtol} $= 2 \times 10^{-4}$ and \id{atol} $= 10^{-8}$ in \eqref{eq:stoppingcriterion}.  The initial iterate was zero ice: $s=b$.

These solver parameters generate a solver with mesh-independent iterations.  Then we fixed the number of degrees of freedom per process at $m_J/P=1281^2 \approx 1.6 \times 10^6$ and solved over $P=1,4,16$ processes.\footnote{Runs were on a single-node workstation with two Intel Xeon 4210 cpus (2.2GHz), a total of 20 physical cores, and 128 GB shared memory.}  The results in Table \ref{tab:results:siaweak} show high-quality weak scaling, that is, nearly-constant run time for mesh refinement under constant $m_J/P$.  The finest mesh result, with $m_J=5121^2=2.6 \times 10^7$ degrees of freedom, needed six minutes to solve for the surface elevation of an ice sheet slightly larger than the current Greenland ice sheet ($1.7$ million $\text{km}^2$) at uniform $350$ m resolution.
\end{example}

@@

>>>>

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python, for $P_1$ and $Q_1$ elements, using the Firedrake library \cite{Rathgeberetal2016}.  Here we show numerical results for the example problems presented in Section \ref{sec:vi}, with a focus on the scaling of FAScd iterations as the mesh resolution increases.  Our primary observations are that each V-cycle dramatically reduces the residual norm, and that very few F-cycles solve most examples to within discretization error.

\begin{example} \label{ex:results:classical}
Consider the unilateral 2D Laplacian obstacle problem, Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\} \subset W^{1,2}(\Omega)$.  We consider two problems, both over square domains $\Omega$, with the very different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$ shown in Figure \ref{fig:results:classical}.  The standard ``ball'' problem has an upper hemisphere obstacle $\psi$ \cite[Chapter 12]{Bueler2021}.  For the ``spiral'' problem \cite[problem 7.1.1]{GraeserKornhuber2009}, $\psi$ is constructed to produce the coincidence set shown.  For this experiment we used a coarsest regular-triangulation mesh with $m_0=5^2$ vertices (32 triangles).  The finest mesh, built by standard refinement, had $m_9=2049^2 \approx 4.2 \times 10^{6}$ vertices.  As our smoother we applied a single RS Newton iteration (Section \ref{sec:implementation}), with three (fixed) IC-preconditioned CG iterations to approximately solve the arising linear systems.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  The initial iterate was set to $w=\max\{0,\psi\}$.

Table \ref{tab:results:classical} shows the resulting iteration counts.  When the RS Newton smoother is applied by itself as a solver (``RS only''), as expected it shows rapidly-increasing iterations, roughly proportional to resolution.  By contrast, V-cycle iterations increase slowly, perhaps logarithmically with the resolution.  For F-cycles we report the number of V-cycles after the ramp (Section \ref{sec:implementation}); a constant number solves the problem.  Comparing the ball and spiral results, the complexity of the coincidence set or free boundary does not appear to strongly influence FAScd performance.  For the spiral results, note that \cite{GraeserKornhuber2009} reports more than 30 iterations of all their tested methods to achieve similar tolerance for $m_J \approx 5\times 10^5$.

Each V-cycle does the work of about eight finest-level preconditioned Krylov (CG+IC) iterations.  Thus for the F-cycle results, where three iterations suffice at all resolutions, we see close to textbook multigrid efficiency\footnote{Textbook multigrid efficiency is imperfectly defined, but consider \cite{BrownSmithAhmadia2013}: ``an order of magnitude reduction in residual per iteration and solution of the fine-level system at a small multiple of the cost of a residual evaluation.''  Work of less than ten residual evaluations is often mentioned as the standard.} despite the low solution regularity.
\end{example}

%\pef{unlike \cite{GraeserKornhuber2009}?}
% Hard to say.  They only look at one resolution, and they don't show something like "ball".  But their iterations are in the 40--80 range; I've noted that.

@@

\begin{example}  \label{ex:results:plap}
We again consider Example \ref{ex:plaplacian} with a unilateral constraint, but now for the nonlinear $p$-Laplacian with both fast ($p=1.5$) and degenerate ($p=4.0$) diffusivities.\footnote{``Fast'' and ``degenerate'' refer to $|\grad u|^{p-2}\to \infty,0$, respectively; ellipticity is lost in the latter case.  Well-posedness and ($p$-dependent) solution regularity for $p$-Laplacian obstacle problems is considered in \cite{ChoeLewis1991}.} We consider the problem in one dimension for ease of visualization. This example reveals the sometimes strong contrast between $V(1,0)$ and $V(0,1)$ results (\pr{up}, \pr{down} $=0$ in Algorithm \ref{alg:fascd}, respectively), and it emphasizes the need for robust smoothers.

Set $\Omega=(-3,3)$, $\psi(x) = -0.2|x|$, $\mathcal{K} = \{v \ge \psi\} \subset W^{1,p}(\Omega)$, and $\ip\ell v = \int_\Omega g v$, where $g(x)=1$ for $|x|<1$ and $g(x)=-1$ for $|x|>1$, in VI \eqref{eq:vi}.  The exact solution is easily computed for any $p$; see Figure \ref{fig:results:plap} and the documentation of the code associated to this paper.  For this experiment we used a coarsest mesh of $m_0=7$ vertices and a finest mesh with $m_9=3073$.  The default smoother uses three (fixed) iterations of RS Newton, with LU solution of the arising linear systems, an $O(m_j)$ smoother in 1D.  The coarse solver is the same but iterated to convergence.  The initial iterate was $w=\psi$, and \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$.  Figure \ref{fig:imagesvcycle} illustrates the major steps of an example V-cycle, showing each of the coarse-correction VI subproblems and solutions.

%\pef{Write down the exact solution?}
% see code reproducibility

@@

Solvers implemented in Firedrake can exploit parallel solvers from the PETSc library \cite{Balayetal2023}, and in fact the above Examples, which were run in serial so as to fix the smoother, also run efficiently in parallel.  The next Examples apply parallel solvers, and Example \ref{ex:results:sia} demonstrates the parallel weak scaling \cite{Bueler2021} of FAScd.

\begin{example}  \label{ex:results:advdiff}
We consider the advection-diffusion VI problem in Example \ref{ex:advectiondiffusion}.  Let $\Omega=(-1,1)^d$ for $d=2,3$, the diffusivity $\eps=0.1$ (equation \eqref{eq:advectiondiffusionstrong}), and we impose zero Dirichlet boundary conditions.  The 2D problem satisfied the coercivity hypothesis with a divergence-free velocity field $\bX = (7+5y,-5x)$, but in 3D, for varieties sake, we set $\bX = (7+5y,-5x,2z)$ so $\grad\cdot\bX \ne 0$.  The discontinuous source term $\phi \in L^\infty(\Omega)$ satisfied $\phi\ge 0$ in half of the domain, with positive circular patches, while values were negative in the other half; details are in the associated code.   While the other Examples in this Section are unilateral, here the solution was box-constrained: $0 \le u \le 1$.  Figure \ref{fig:results:advdiff} shows coincidence sets for the 2D problem; the 3D sets are comparable in regularity and general pattern.

@@

F-cycle iteration results are shown in Table \ref{tab:results:advdiff}.  Despite the hint of growing iterations under mesh refinement, a few iterations sufficed up to memory-limited resolutions.  For any of the higher resolution runs, the FAScd F-cycle solver took less time than generating the $P_1$ mesh hierarchy.  The local Péclet number $\text{Pe}=\|\bX(x,y,z)\|_2 L \eps^{-1}$, with $L=2$ the width of $\Omega$, varied from 40 to 260; the mesh Péclet number $m_J^{-1/d}\,\text{Pe}$ varied accordingly.  Thus the $m_J\le 121^d$ problems were advection-dominated while those with $m_J\ge 481^d$ were diffusion-dominated, but the smoother was the same in all cases, and the iterations were nearly constant.
\end{example}

%\pef{I would have said you need $Pe < 1$ to be diffusion-dominated, not have $Pe = 40$?}
% Clarified that this is a mesh Peclet thing.

% see git@bitbucket.org:pefarrell/fascd.git: examples/pollutant.py
% and https://github.com/bueler/mcd-extended: paper/results/pollutant.sh|txt
\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}c@{\hskip 4mm}c}
\emph{levels} & $m_J$ & $d=2$ & $d=3$ \\ \cmidrule{1-4}
 1 &    $16^d$ & 1 & 1 \\
 2 &    $31^d$ & 2 & 2 \\
 3 &    $61^d$ & 2 & 2 \\
 4 &   $121^d$ & 2 & 2 \\
 5 &   $241^d$ & 2 & 3 \\
 6 &   $481^d$ & 2 \\
 7 &   $961^d$ & 3 \\
 8 &  $1921^d$ & 3 \\
 9 &  $3841^d$ & 3
\end{tabular}
\bigskip
\caption{F-cycle iterations for 2D ($P_1$, serial) and 3D ($Q_1$, 16 processes) solutions of the advection-diffusion bi-obstacle VI problem in Example \ref{ex:results:advdiff}.}
\label{tab:results:advdiff}
\end{table}

\begin{example}   \label{ex:results:sia}
We consider the steady shallow ice approximation (SIA) obstacle problem presented in Example \ref{ex:sia}, in which the surface elevation is constrained to exceed the bedrock elevation, $s\ge b$.  We set $\Omega=(0,L)^2$ with $L=1800$ km.  The data consist of a radial surface mass balance function $a$ \cite[equation (5.122)]{GreveBlatter2009} and a bumpy, synthetic map $b\in C^1(\Omega)$, with an elevation range of $1800$ m; details for $b$ are in the associated code.  A numerical solution is shown in Figure \ref{fig:results:siascene}, colored by the solution surface flow speed, i.e.~the magnitude of $\mathbf{u} = (n+1)(n+2)^{-1} \Gamma (s-b)^{n+1} |\grad s|^{n-1} \grad s$.  As noted in Example \ref{ex:sia}, and seen in this Figure, $|\grad s|$ is unbounded at the free boundary; the solution $s$, while continuous, has very low regularity.

@@

We applied FAScd F-cycles using $Q_1$ elements and a smoother of four (fixed) iterations of RS Newton (Section \ref{sec:implementation}), with quadratic backtracking line search.  The linear step equations were approximately solved, in parallel, by three (fixed) iterations of preconditioned GMRES and a ASM+ILU preconditioner.  The coarse solver was RS+LU, applied redundantly over all processors.  The coarse mesh was relatively fine, with $m_0=21^2$ vertices and 400 cells.  Because of the low regularity of the solution we used \id{rtol} $= 2 \times 10^{-4}$ and \id{atol} $= 10^{-8}$ in \eqref{eq:stoppingcriterion}.  The initial iterate was zero ice: $s=b$.  These parameters generated a solver with mesh-independent iterations.

For a weak-scaling study we fixed the number of degrees of freedom per process at $m_J/P=1281^2 \approx 1.6 \times 10^6$ and solved over $P=1,4,16$ processes.\footnote{Runs were on a single-node workstation with two Intel Xeon 4210 cpus (2.2GHz), a total of 20 physical cores, and 128 GB shared memory.}  The results in Table \ref{tab:results:siaweak} show nearly-constant run time for this parallel refinement path.  The finest mesh result, with $m_J=5121^2=2.6 \times 10^7$ degrees of freedom, needed six minutes to solve for the steady surface elevation of an ice sheet larger than the current Greenland ice sheet ($1.7$ million $\text{km}^2$) at uniform $350$ m resolution.
\end{example}

@@

\subsection*{Code availablility} \label{sec:code}  For reproducibility we have archived the exact software versions used to produce the results in this paper, including the correct Firedrake version, at FIXME Zenodo?

\subsection*{Acknowledgements} \label{sec:acknowledgements}  EB was supported by a Faculty Development Travel Award from United Academics, University of Alaska Fairbanks.  FIXME PF grant support?
